---
title: "Solving Perfect Multicollinearity with Nested Geographic Variables in R"
author: "Paul Mosquera"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    theme: flatly
    highlight: tango
    number_sections: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = "center")
```

```{r libraries, include=FALSE}
library(tidyverse)
library(car)
library(here)
library(kableExtra)
library(plotly)
```

# The Problem: Nested Geographic Variables in Regression

Imagine you are building a model to explain housing prices across municipalities in Belgium. Your dataset contains two geographic variables: the administrative region each municipality belongs to (AdminArea), and its province (Province). Naturally, you want to include both because regions capture broad geographic effects and provinces capture finer local differences.

The problem is that these two variables are not independent. Every province belongs to exactly one region, always. Hainaut is always in Waals_Gewest. Limburg is always in Vlaams_Gewest. There are no exceptions. This rigid hierarchical structure means that once you know the province of a municipality, you already know its region with 100% certainty.

When you try to include both variables in a regression model, R cannot distinguish the effect of the region from the effect of the province. The model literally cannot be estimated. This is called perfect multicollinearity.

## The Mathematical Reason

In linear regression, the coefficients are estimated using the formula:

$$\hat{\beta} = (X'X)^{-1}X'y$$

where $X$ is the design matrix containing all predictor variables. For this formula to work, the matrix $X'X$ must be invertible, which requires that no column of $X$ can be written as an exact linear combination of other columns.

When Province is nested within AdminArea, the region column in $X$ is completely determined by the province columns. Concretely:

$$\text{AdminArea}_i = \begin{cases} \text{Vlaams\_Gewest} & \text{if } \text{Province}_i \in \{\text{Antwerpen, Limburg, Oost\_Vlaanderen, ...}\} \\ \text{Waals\_Gewest} & \text{if } \text{Province}_i \in \{\text{Hainaut, Liege, Luxembourg, ...}\} \end{cases}$$

This means one column of $X$ is a perfect linear combination of others, which makes the determinant of $X'X$ equal to zero:

$$\det(X'X) = 0 \implies (X'X)^{-1} \text{ does not exist}$$

If you sum the province dummy columns belonging to Vlaams_Gewest, you get exactly the AdminArea dummy column:

$$x_{\text{Antwerpen}} + x_{\text{Limburg}} + x_{\text{Oost\_Vl}} + \cdots = x_{\text{Vlaams\_Gewest}}$$

The design matrix $X$ looks like this conceptually:

```{r matrix-table, echo=FALSE}
matrix_example <- data.frame(
  Municipality  = c("Kortenaken", "Hamoir", "Hasselt"),
  Vlaams_Gewest = c(1, 0, 1),
  Hainaut       = c(0, 0, 0),
  Liege         = c(0, 1, 0),
  Limburg       = c(0, 0, 1),
  `...`         = c("...", "...", "...")
)

kable(matrix_example, align = "c",
      caption = "Conceptual design matrix X showing the redundancy of AdminArea") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

The AdminArea column is redundant. It carries no information that the province columns do not already contain. The matrix is singular and the model collapses.

## The Dataset

We use housing price data from Belgium with 566 municipalities, focusing on two administrative regions: Vlaams_Gewest (Flemish Region) and Waals_Gewest (Walloon Region). Each municipality belongs to exactly one province, which belongs to exactly one region.

```{r load-data}
df <- read.table(here("data", "HousePrices.txt"), header = TRUE)

df <- df |> filter(AdminArea %in% c("Vlaams_Gewest", "Waals_Gewest"))

df$AdminArea <- factor(df$AdminArea, levels = c("Vlaams_Gewest", "Waals_Gewest"))
df$Province  <- droplevels(factor(df$Province))

cat("Rows after filter:", nrow(df), "\n")
cat("AdminArea levels:", levels(df$AdminArea), "\n")
cat("Province levels:", levels(df$Province), "\n")
```

## The Hierarchical Structure

Before attempting any model, it helps to visualize exactly how provinces are nested within regions. This makes the source of the problem immediately obvious.

```{r hierarchy-plot}
hierarchy <- df |>
  count(AdminArea, Province) |>
  arrange(AdminArea, desc(n))

p1 <- ggplot(hierarchy,
             aes(x = reorder(Province, n), y = n, fill = AdminArea,
                 text = paste("Province:", Province,
                              "<br>Region:", AdminArea,
                              "<br>Municipalities:", n))) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("Vlaams_Gewest" = "#2196F3",
                                "Waals_Gewest"  = "#FF9800")) +
  labs(title = "Municipalities per Province colored by Region",
       x = NULL, y = "Number of municipalities", fill = "Region") +
  theme_minimal(base_size = 12)

ggplotly(p1, tooltip = "text")
```

The color pattern is the key insight: every province bar is a single color. There is no province that is split between two regions. This perfect separation is precisely what causes the model to fail.

## Demonstrating the Failure

```{r demonstrate-failure}
model_naive <- lm(med_price_house ~ AdminArea + Province, data = df)
summary(model_naive)
```

Look carefully at the output. Several province coefficients appear as NA. This is R silently telling you that it cannot estimate those coefficients because the information they would carry is perfectly confounded by the AdminArea variable. The model is broken before it even starts.

# Diagnosis: Confirming the Problem

## Variance Inflation Factor (VIF)

The VIF measures how much the variance of a coefficient estimate is inflated due to its correlation with other predictors:

$$VIF_j = \frac{1}{1 - R^2_j}$$

| VIF Range | Interpretation |
|---|---|
| VIF = 1 | No correlation with other predictors |
| VIF between 1 and 5 | Generally acceptable |
| VIF above 10 | Serious concern |
| VIF approaching infinity | Perfect multicollinearity |

```{r vif-naive}
tryCatch(
  car::vif(model_naive),
  error = function(e) cat("VIF Error:", conditionMessage(e), "\n")
)
```

## Rank of the Design Matrix

```{r rank-check}
X_naive <- model.matrix(~ AdminArea + Province, data = df)

cat("Number of columns in X:", ncol(X_naive), "\n")
cat("Rank of X:", qr(X_naive)$rank, "\n")
cat("Rank deficiency:", ncol(X_naive) - qr(X_naive)$rank, "\n")
```

The rank deficiency tells you exactly how many columns are redundant. One column can be expressed as a perfect linear combination of the others, confirming that AdminArea and Province cannot coexist in the same model without a structural fix.

# Statistical Assumptions Check

Before applying the solution, we verify whether the conditions for classical linear regression are reasonably met in this dataset. The goal here is not to validate a final predictive model but to establish whether this methodological solution is applicable in this context.

It is important to clarify the modeling objective upfront: we are not building the most accurate model possible for Belgian housing prices. We are using a deliberately simple model with three predictors to demonstrate when and how this solution applies. This distinction matters for interpreting the assumption checks that follow.

## Normality of the Response Variable

Classical linear regression does not strictly require normality of the response variable. What matters is that the residuals are approximately normally distributed. However, inspecting the response distribution gives us useful information about potential issues.

```{r response-distribution}
df$log_price <- log(df$med_price_house)

p_hist <- ggplot(df, aes(x = med_price_house)) +
  geom_histogram(bins = 40, fill = "#2196F3", color = "white", alpha = 0.8) +
  labs(title = "Distribution of med_price_house (original scale)",
       x = "med_price_house", y = "Frequency") +
  theme_minimal(base_size = 12)

p_hist_log <- ggplot(df, aes(x = log_price)) +
  geom_histogram(bins = 40, fill = "#FF9800", color = "white", alpha = 0.8) +
  labs(title = "Distribution of log(med_price_house)",
       x = "log(med_price_house)", y = "Frequency") +
  theme_minimal(base_size = 12)

ggplotly(p_hist)
ggplotly(p_hist_log)
```

```{r qqplot-response}
p_qq <- ggplot(df, aes(sample = med_price_house)) +
  stat_qq(color = "#2196F3") +
  stat_qq_line(color = "gray40") +
  labs(title = "QQ-plot: med_price_house (original scale)") +
  theme_minimal(base_size = 12)

p_qq_log <- ggplot(df, aes(sample = log_price)) +
  stat_qq(color = "#FF9800") +
  stat_qq_line(color = "gray40") +
  labs(title = "QQ-plot: log(med_price_house)") +
  theme_minimal(base_size = 12)

p_qq
p_qq_log
```

The original scale shows moderate right skewness. The log transformation improves symmetry noticeably. Despite this, we keep the response in its original scale for a deliberate reason: our objective is to demonstrate the methodological solution with directly interpretable coefficients in monetary units. A coefficient of "Hainaut has prices 32 thousand below the global average" is immediately meaningful to any stakeholder. Under log transformation, coefficients would represent approximate percentage deviations, which are harder to communicate in applied contexts.

With a sample of 566 observations, the Central Limit Theorem provides sufficient protection for the sampling distribution of the estimators to be approximately normal even when the response is moderately skewed. This makes OLS inference reasonably valid despite the departure from normality.

## Homoscedasticity Across Geographic Groups

This is the most critical assumption to verify for our specific solution. Since we are including two geographic levels in the same model, we need to assess whether the variance of housing prices is approximately constant across regions and provinces.

We use Levene's test, which is robust to departures from normality and tests whether group variances are equal.

```{r levene-region}
cat("Levene's test for homoscedasticity across AdminArea:\n")
leveneTest(med_price_house ~ AdminArea, data = df)
```

```{r levene-province}
cat("Levene's test for homoscedasticity across Province:\n")
leveneTest(med_price_house ~ Province, data = df)
```

```{r variance-by-group}
var_by_region <- df |>
  group_by(AdminArea) |>
  summarise(
    n        = n(),
    mean     = round(mean(med_price_house), 2),
    sd       = round(sd(med_price_house), 2),
    variance = round(var(med_price_house), 2)
  )

kable(var_by_region,
      caption = "Descriptive statistics of med_price_house by AdminArea") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

```{r variance-boxplot}
p_box <- ggplot(df, aes(x = AdminArea, y = med_price_house, fill = AdminArea,
                         text = paste("Region:", AdminArea,
                                      "<br>Price:", med_price_house))) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("Vlaams_Gewest" = "#2196F3",
                                "Waals_Gewest"  = "#FF9800")) +
  labs(title = "Housing price distribution by AdminArea",
       x = NULL, y = "med_price_house") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")

ggplotly(p_box, tooltip = "text")
```

If Levene's test rejects homoscedasticity, this does not invalidate the solution but it does inform how we interpret the results. OLS estimators remain unbiased under heteroscedasticity. What changes is that standard errors may be slightly underestimated, which affects the precision of hypothesis tests. For the purposes of this methodological demonstration, where the focus is on the structural fix rather than formal inference, this is an acceptable limitation.

If formal inference were the primary goal, the appropriate response would be to either use heteroscedasticity-consistent standard errors (HC standard errors via the `sandwich` package) or consider a multilevel model with `lme4` that explicitly models group-level variance.

# The Solution

The fix has two steps. First, we create a BASE reference category by merging one representative province from each region. Second, we apply sum-to-zero contrasts to both AdminArea and the recoded province variable. Together, these two steps break the perfect linear dependence while keeping both geographic levels in the model.

## Step 1: Choosing the BASE Province

Instead of arbitrarily picking a province to merge into BASE, we select it using a principled criterion: the province with the most observations in each region. This choice maximizes the stability of the reference category because it is backed by the most data.

```{r choose-base}
base_provinces <- df |>
  count(AdminArea, Province) |>
  group_by(AdminArea) |>
  slice_max(n, n = 1) |>
  ungroup()

kable(base_provinces,
      caption = "Selected BASE province per region based on highest observation count") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

```{r create-province2}
df$Province2 <- as.character(df$Province)

for (i in seq_len(nrow(base_provinces))) {
  area <- base_provinces$AdminArea[i]
  prov <- base_provinces$Province[i]
  df$Province2[df$AdminArea == area & df$Province == prov] <- "BASE"
}

lv <- c(setdiff(unique(df$Province2), "BASE"), "BASE")
df$Province2 <- factor(df$Province2, levels = lv)

df |> count(Province2) |> arrange(desc(n)) |>
  kable(caption = "Municipality count per Province2 after BASE recoding") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

Notice that BASE now groups the most representative province from each region into a single reference category. This is what breaks the perfect linear dependence: BASE contains municipalities from both regions, so Province2 is no longer a perfect predictor of AdminArea.

## Why Does This Work?

After the recoding, the BASE category contains municipalities from both Vlaams_Gewest and Waals_Gewest, which means:

$$\text{Province2}_i = \text{BASE} \not\Rightarrow \text{AdminArea}_i \text{ is determined}$$

The perfect linear dependence is broken. The design matrix now has full rank and the model can be estimated.

```{r visualize-base}
hierarchy2 <- df |>
  count(AdminArea, Province2) |>
  arrange(AdminArea, desc(n))

p2 <- ggplot(hierarchy2,
             aes(x = reorder(Province2, n), y = n, fill = AdminArea,
                 text = paste("Province2:", Province2,
                              "<br>Region:", AdminArea,
                              "<br>Municipalities:", n))) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("Vlaams_Gewest" = "#2196F3",
                                "Waals_Gewest"  = "#FF9800")) +
  labs(title = "Municipalities per Province2 after BASE recoding",
       x = NULL, y = "Number of municipalities", fill = "Region") +
  theme_minimal(base_size = 12)

ggplotly(p2, tooltip = "text")
```

Compare this plot with the one in Section 1. The BASE bar is now split between two colors, which visually confirms that the perfect separation has been broken.

## Step 2: Applying Sum-to-Zero Contrasts

Sum-to-zero contrasts, also known as deviation coding, make each coefficient represent the deviation of that category from the overall mean rather than the deviation from a single reference category:

$$\sum_{k=1}^{K} \beta_k = 0$$

The last level, BASE in our case, is the implicit reference and its coefficient is recovered as minus the sum of all others. This makes every coefficient directly interpretable as a deviation from the global average, which is especially useful when both geographic levels are in the model simultaneously.

```{r apply-contrasts}
kA <- nlevels(df$AdminArea)
kP <- nlevels(df$Province2)

CA <- contr.sum(kA)
CP <- contr.sum(kP)

colnames(CA) <- levels(df$AdminArea)[1:(kA - 1)]
colnames(CP) <- levels(df$Province2)[1:(kP - 1)]

contrasts(df$AdminArea) <- CA
contrasts(df$Province2) <- CP

cat("AdminArea contrasts:\n")
contrasts(df$AdminArea)

cat("\nProvince2 contrasts:\n")
contrasts(df$Province2)
```

## Fitting the Model

With the recoding and contrasts in place, we fit a deliberately simple model that includes both geographic levels simultaneously along with mean income as a continuous predictor. The simplicity is intentional: the goal is to demonstrate the methodological solution clearly, not to build an exhaustive predictive model.

```{r fit-model}
df$Mean_income_c <- df$Mean_income - mean(df$Mean_income, na.rm = TRUE)

model_solution <- lm(med_price_house ~ AdminArea + Province2 + Mean_income_c,
                     data = df)

summary(model_solution)
```

# Verification: Did the Solution Work?

## VIF After the Fix

```{r vif-solution}
vif_solution <- car::vif(model_solution)

vif_solution |>
  as.data.frame() |>
  kable(caption = "GVIF values for the fixed model", digits = 4) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

## Rank of the New Design Matrix

```{r rank-check-solution}
X_solution <- model.matrix(~ AdminArea + Province2 + Mean_income_c, data = df)

cat("Number of columns in X:", ncol(X_solution), "\n")
cat("Rank of X:", qr(X_solution)$rank, "\n")
cat("Rank deficiency:", ncol(X_solution) - qr(X_solution)$rank, "\n")
```

A rank deficiency of zero confirms that the design matrix now has full column rank. Every coefficient can be estimated without ambiguity.

## Visual Comparison of VIF Before and After

```{r vif-comparison-plot}
model_partial <- lm(med_price_house ~ Province + Mean_income_c, data = df)

vif_before <- car::vif(model_partial)
vif_after  <- car::vif(model_solution)

vif_before_df <- data.frame(
  variable = names(vif_before[, 3]),
  vif      = vif_before[, 3],
  model    = "Before fix"
)

vif_after_df <- data.frame(
  variable = names(vif_after[, 3]),
  vif      = vif_after[, 3],
  model    = "After fix"
)

vif_combined <- bind_rows(vif_before_df, vif_after_df)

p3 <- ggplot(vif_combined,
             aes(x = reorder(variable, vif), y = vif, fill = model,
                 text = paste("Variable:", variable,
                              "<br>GVIF:", round(vif, 3),
                              "<br>Model:", model))) +
  geom_col(position = "dodge") +
  geom_hline(yintercept = 2, linetype = "dashed", color = "gray40") +
  coord_flip() +
  scale_fill_manual(values = c("Before fix" = "#EF5350",
                                "After fix"  = "#42A5F5")) +
  labs(title = "VIF Comparison Before and After the Fix",
       x = NULL, y = "GVIF^(1/(2*Df))", fill = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")

ggplotly(p3, tooltip = "text")
```

## Model Diagnostics

Beyond VIF, we verify the classical regression assumptions on the fitted model. These diagnostics do not validate the solution itself, which was already confirmed by the rank analysis and VIF, but they tell us how much we can trust the coefficient estimates and standard errors for inference.

```{r model-diagnostics}
par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
plot(model_solution, which = c(1, 2, 3, 5))
par(mfrow = c(1, 1))
```

```{r residuals-by-group}
df$residuals <- residuals(model_solution)

p_resid <- ggplot(df, aes(x = AdminArea, y = residuals, fill = AdminArea,
                           text = paste("Region:", AdminArea,
                                        "<br>Residual:", round(residuals, 2)))) +
  geom_boxplot(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  scale_fill_manual(values = c("Vlaams_Gewest" = "#2196F3",
                                "Waals_Gewest"  = "#FF9800")) +
  labs(title = "Residuals by AdminArea",
       subtitle = "Systematic patterns here would indicate model misspecification",
       x = NULL, y = "Residuals") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")

ggplotly(p_resid, tooltip = "text")
```

If the residual spread differs substantially between regions, this confirms the heteroscedasticity detected by Levene's test. In that case, the coefficient estimates remain unbiased but standard errors should be interpreted with caution. For a methodological demonstration like this one, that is acceptable. For formal inference in a production setting, heteroscedasticity-consistent standard errors or a multilevel model would be more appropriate.

# Interpretation: Reading Coefficients Under Sum-to-Zero Contrasts

## The Intercept

Under sum-to-zero contrasts, the intercept represents the overall mean of the response variable when all continuous predictors are at their mean values. It is not the mean of a specific reference category as it would be under treatment coding.

```{r intercept}
intercept_val <- round(coef(model_solution)["(Intercept)"], 2)
cat("Intercept (global mean housing price):", intercept_val, "thousands\n")
```

## Region and Province Coefficients

Each coefficient represents the deviation of that category from the overall mean. The implicit BASE coefficient is recovered as:

$$\beta_{\text{BASE}} = -\sum_{k \neq \text{BASE}} \beta_k$$

```{r recover-base}
province_coefs <- coef(model_solution)[grep("Province2", names(coef(model_solution)))]
beta_base <- -sum(province_coefs)
cat("Implicit BASE coefficient:", round(beta_base, 4), "\n")
```

## Comparing Two Provinces Directly

$$\hat{y}_A - \hat{y}_B = \beta_A - \beta_B$$

```{r province-comparison}
all_province_coefs <- c(province_coefs, BASE = beta_base)

comparison <- data.frame(
  Province    = gsub("Province2", "", names(all_province_coefs)),
  Coefficient = round(all_province_coefs, 2)
) |> arrange(desc(Coefficient))

kable(comparison,
      caption = "Province deviations from the global mean housing price",
      row.names = FALSE) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) |>
  row_spec(which(comparison$Coefficient > 0), color = "#1565C0") |>
  row_spec(which(comparison$Coefficient < 0), color = "#B71C1C")
```

```{r comparison-plot}
p4 <- ggplot(comparison,
             aes(x = reorder(Province, Coefficient),
                 y = Coefficient,
                 fill = Coefficient > 0,
                 text = paste("Province:", Province,
                              "<br>Deviation:", Coefficient, "thousands"))) +
  geom_col() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "#42A5F5", "FALSE" = "#EF5350"),
                    labels = c("TRUE" = "Above average",
                               "FALSE" = "Below average")) +
  labs(title = "Province Deviations from the Overall Mean Housing Price",
       x = NULL, y = "Deviation from global mean (thousands)", fill = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")

ggplotly(p4, tooltip = "text")
```

## A Concrete Example

```{r concrete-example}
beta_vb <- coef(model_solution)["Province2Vlaams_Brabant"]
beta_ha <- coef(model_solution)["Province2Hainaut"]
diff    <- beta_vb - beta_ha

cat("Vlaams_Brabant coefficient:", round(beta_vb, 2), "\n")
cat("Hainaut coefficient:", round(beta_ha, 2), "\n")
cat("Expected price difference (Vlaams_Brabant vs Hainaut):",
    round(diff, 2), "thousands\n")
```

A municipality in Vlaams_Brabant is expected to have a housing price approximately that many thousands higher than a comparable municipality in Hainaut, holding mean income constant. This is a direct and clean comparison that sum-to-zero contrasts make possible without any additional calculation.

# Applicability and Limitations

## When This Solution Is Appropriate

This solution is appropriate when all of the following conditions hold. The analyst needs both hierarchical levels in the model simultaneously for interpretability purposes. The sample size is large enough for the Central Limit Theorem to protect inference under moderate violations of normality. The homoscedasticity assumption is at least approximately met, or formal inference is not the primary goal. The modeling framework is classical linear regression rather than a formal multilevel analysis.

## When to Consider Alternatives

If the primary goal is to model within-group correlation correctly, a multilevel model with `lme4` is more appropriate. If heteroscedasticity across groups is severe, heteroscedasticity-consistent standard errors via the `sandwich` package should be used alongside this solution. If the response variable requires transformation for valid inference, the log-transformed response can be used with the understanding that coefficients will represent approximate percentage deviations rather than absolute differences.

## Common Scenarios

Latin America is particularly rich in nested geographic structures. In Ecuador, parishes are nested within cantons, cantons within provinces, and provinces within regions (Costa, Sierra, Amazonia, Insular). The same applies to municipalities nested within departments in Colombia and Peru, and to districts nested within cities in any urban study. Beyond geography, the same structure appears in education research when schools are nested within districts, in clinical studies when patients are nested within hospitals, and in business analytics when products are nested within categories.

## The General Recipe

```{r recipe-table, echo=FALSE}
recipe <- data.frame(
  Step = 1:7,
  Action = c(
    "Identify the two hierarchical levels that cause the problem",
    "Check homoscedasticity across groups with Levene's test",
    "Choose a BASE category using a principled criterion such as the most frequent group at the lower level within each upper level group",
    "Recode the lower level variable replacing the chosen categories with BASE",
    "Set BASE as the last factor level",
    "Apply contr.sum to both variables",
    "Verify with VIF and rank analysis that the deficiency is resolved"
  )
)

kable(recipe,
      caption = "General recipe for solving perfect multicollinearity with nested geographic variables") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE)
```

## Summary

This document demonstrates a methodological solution for a specific and recurring problem in applied regression: perfect multicollinearity caused by nested geographic variables. The solution is statistically valid under classical linear regression assumptions and is especially valuable when interpretability of fixed coefficients at both geographic levels is the primary goal.

The approach is not a replacement for multilevel models. It is a practical and accessible alternative when the analyst works within a classical regression framework and needs both geographic levels in the model with directly interpretable coefficients. Understanding when to apply it, and when its assumptions are sufficiently met, is the core statistical competency this document aims to build.

This solution was developed during the course Applied Statistical Models I at ESPOL, Ecuador, as part of a group project. The methodological contribution documented here was developed independently by the me.